---
title: "Working with Bchron Package"
output: html_notebook
---

Trial for running Bchron
I've already calibrated my radiocarbon dates with Calib, but I wanted to mess around with the Bchron package mostly to use the BchronDensity function to visualize my dates. I'm using this site: https://cran.r-project.org/web/packages/Bchron/vignettes/Bchron.html for reference.

```{r packages}
library(Bchron)
library(tidyverse)
library(ggplot2)
library(here)
library(cowplot)
theme_set(theme_cowplot())
```

Calibrating radiocabon dates
```{r Testing Calibration}
red_c14 <- read.csv(here("red14_bchron.csv"))
# testing for first date
# age1 = BchronCalibrate(ages = 480, ageSds = 30, #standard deviation of age calCurves = "intcal13", #calibration curve (choice of intcal13, shcal13 and marine13) ids = "WOMT-01d5") #id number of date
# summary(age1)
# plot(age1, xlab = "Age (years BP)")
```

```{r eval=FALSE, include=FALSE}
red_cal <- rep("intcal13", times = length(red_c14$ages))
all_ages = BchronCalibrate(ages = red_c14$ages,
                           ageSds = red_c14$SD,
                           positions = red_c14$depth,
                           calCurves = red_cal)
summary(all_ages)
plot(all_ages, withPositions=TRUE, 
     xlab = "Age (cal years BP)",
     ylab = "Depth (cm)")
```

Alright, now let's mess with the BchronDensity function:
```{r BchronDensity}
# removing 6k dates
red_c14 <- read.csv(here("red14_bchron.csv"))
red_cal <- rep("intcal13", times = length(red_c14$ages))
red_density = BchronDensity(ages = red_c14$ages,
                            ageSds = red_c14$SD,
                            calCurves = red_cal)
 plot(red_density)
# rm(red_density)
```

4/18 - I'm going to rerun the prior two graphs, but remove the two abnormal sites (GOV_01 and WORM_03)
```{r}
# removed the sites from the red_c14 file in excel for time's sake
red_c14_normal <- read.csv(file = "red14_bchron_normal.csv")

# calibrating
red_ages_norm <- red_c14_normal$ages
red_SDs_norm <- red_c14_normal$SD
red_depth_norm <- red_c14_normal$depth
red_cal_norm <- rep("intcal13", times = length(red_ages_norm))

all_ages_norm = BchronCalibrate(ages = red_ages_norm,
                           ageSds = red_SDs_norm,
                           positions = red_depth_norm,
                           calCurves = red_cal_norm)
summary(all_ages_norm)
plot(all_ages_norm, withPositions=TRUE)
```


```{r}
# bchron density
# red_density_norm = BchronDensity(ages = red_ages_norm, ageSds = red_SDs_norm, calCurves = red_cal_norm)
# plot(red_density_norm, bty = "n", xlab = "Age (cal years BP)")
rm(red_density_norm)
```

but, what if I could do this in ggplot...

```{r}
# install.packages("devtools")
library(devtools)
devtools::install_github('andrewcparnell/Bchron')
library(Bchron)
data(red_c14_normal)
red_dens <- BchronDensity(ages=  red_c14_normal$ages, ageSds = red_c14_normal$SD, calCurves = red_cal_norm, numMix = 50, store_density = TRUE)
plot(red_dens$ageGrid, red_dens$densities, type = 'l', bty = "n")  
plot(red_dens, bty = "n", 
     xlab = "Ages (cal year BP)")
```

I should probably also do this for the colluvial hollow/etc
```{r}
# using just the non soil sites
old_red_input <- red_c14[which(red_c14$ages >= 4000),]
old_red_cal_curve <- rep("intcal13", times = length(old_red_input$ages))
old_red_c14 <- BchronCalibrate(ages = old_red_input$ages, # calibrating
                                   ageSds = old_red_input$SD,
                                   positions = old_red_input$depth,
                                   calCurves = old_red_cal_curve)
# plotting by position
plot(old_red_c14, withPositions = TRUE)

# plotting by density
red_old_dens <- BchronDensity(ages= old_red_input$ages,
                          ageSds = old_red_input$SD,
                          calCurves = old_red_cal_curve, numMix = 50,
                          store_density = TRUE)

plot(red_old_dens$ageGrid, red_old_dens$densities, type = 'l', bty = "n")  
plot(red_old_dens, bty = "n", 
     xlab = "Age (cal year BP)")
```
Alright, let's get that out of Bchrondensity format:
```{r}
# need to change out of BchronDensity format
red_dens_ggplot <- data.frame(red_dens$ageGrid, red_dens$densities) # younger dates
colnames(red_dens_ggplot) <- c("age_grid", "densities")
red_dens_ggplot$type <- rep("ms", each = length(red_dens_ggplot$age_grid))
old_dens_ggplot <- data.frame(red_old_dens$ageGrid, red_old_dens$densities) # older dates
colnames(old_dens_ggplot) <- c("age_grid", "densities")
old_dens_ggplot$type <- rep("old", each = length(old_dens_ggplot$age_grid))

dens_ggplot <- rbind(red_dens_ggplot, old_dens_ggplot)

# creating plots
old_dates_plot <- ggplot(old_dens_ggplot, aes(x = old_dens_ggplot$red_old_dens.ageGrid, y = old_dens_ggplot$red_old_dens.densities)) + geom_line() + background_grid(major = "xy") + labs(x = " ", y = " ") 
old_dates_plot
plot_grid(soil_dates_plot, old_dates_plot, labels = c("A.", "B.")) + geom_point(data = dates, aes(x = dates$median_cal_age, y = dates$y_axis, color = "red", shape = 4))


# let's try this differently
ggplot(dens_ggplot, aes(x = dens_ggplot$age_grid, dens_ggplot$densities)) + geom_line() + facet_wrap(~dens_ggplot$type)
```



```{r}
# i might try the facet_grid function:
# need to change out of BchronDensity format, and want to clean up my names a bit
    # younger dates first
    red_dens_ggplot <- data.frame(red_dens$ageGrid, red_dens$densities) 
    red_dens_ggplot$type <- rep(1, length.out = length(red_dens_ggplot$red_dens.ageGrid))
    names(red_dens_ggplot) <- c("age_grid", "densities", "type")
    red_dens_ggplot
    # same thing for older
    old_dens_ggplot <- data.frame(red_old_dens$ageGrid, red_old_dens$densities) 
    old_dens_ggplot$type <- rep(2,
                                length.out =length(old_dens_ggplot$red_old_dens.ageGrid))
    names(old_dens_ggplot) <- c("age_grid", "densities", "type")
# so, let's put them together
complete_density <- rbind(red_dens_ggplot, old_dens_ggplot)

# now, to plot:
complete_density$type2 <- factor(complete_density$type, labels = c("Mineral Soils", "Colluvial Hollow")) # the solution for my facet label problem, no idea why

ggplot(complete_density, aes(x = complete_density$age_grid, y = complete_density$densities)) + geom_line() + facet_grid(. ~ complete_density$type2, scales = "free_x") + background_grid(major = "xy") + labs(x = "Age (cal year BP)", y = "Densities")

```
Pretty happy with that, minus the weird overlap. I think I may just clip the agegrid for the younger dates, I think that may solve the issue. 
```{r}
# clipping overlap
red_dens_ggplot <- red_dens_ggplot[-which(red_dens_ggplot$age_grid > 4500),] 
old_dens_ggplot <- old_dens_ggplot[-which(old_dens_ggplot$age_grid < 4250),]

complete_density <- rbind(red_dens_ggplot, old_dens_ggplot)
complete_density$type2 <- factor(complete_density$type, labels = c("Mineral Soils", "Colluvial Hollow/Alluvial Fan"))

# giving dates a y-axis
dates <- read.csv("dates.csv")
dates <- dates[-which(is.na(dates$error)),]
length(dates$median_cal_age)
dates$y_axis <- rep(0.003, each = 40)

density_facet_plot <- ggplot(complete_density, 
                             aes(x = complete_density$age_grid, y = complete_density$densities)) +  geom_line() + 
  facet_grid(. ~ complete_density$type2, scales = "free_x") + 
  background_grid(major = "xy") + 
  labs(x = "Age (cal year BP)", y = "Densities",
       title = "Estimate of Site Activity", 
       subtitle = "Results of Bchron Gaussian Mixture Model") + 
  panel_border() +
  theme(plot.title = element_text(hjust = 0)) 
  
density_facet_plot + geom_point(data = dates, aes(x = dates$median_cal_age, y = dates$y_axis, color = "red", shape = 4))
ggplot(dates, aes(x = dates$median_cal_age, y = dates$y_axis)) + geom_point()

density_facet_plot

save_plot("density_facet_plot.png", density_facet_plot, ncol = 2, base_aspect_ratio = 1.1 )
```
Yeah!!

# Related radiocarbon figures
Now that I've tackled the Bchron figures, I should also probably back up and tackle some of the more straightforward radiocarbon graphs (date vs depth, age reversals, etc). I'll start by checking on what sorts of data files I have for that stuff, and how I'll need to clean it up to make it work. 
```{r}
# plotting
ggplot(red_c14, aes(x = red_c14$median_cal_age, y = red_c14$depth, color =
                      factor(red_c14$site))) + 
  geom_point() + facet_grid(.~red_c14$site_type, scales = "fixed")  +
  background_grid(major="xy") + scale_y_reverse()  + panel_border()
```


```{r}
# ggplot(dates, aes(x = dates$median_cal_age, y = dates$depth, color = factor(dates$site_type))) + geom_point() + facet_grid(dates$site_type ~ dates$site, scales = "fixed")  + background_grid(major="xy") + scale_y_reverse()  + panel_border()
```
Well, that's a nightmare. I think the best way to clean that up is to split data between mineral soil and the other two (as I did with the bchron data)
```{r}
ggplot(red_c14, aes(x = red_c14$median_cal_age, y = red_c14$depth, color = red_c14$site_type, fill = red_c14$site_type, shape = red_c14$site_type)) + 
  geom_point(size = 4)

age_date_plot <- ggplot(red_c14, aes(x = red_c14$med_Cal_age, y = red_c14$depth, 
                    color = red_c14$site_type, fill = red_c14$site_type, 
                    shape = red_c14$site_type)) + 
  geom_point(size = 4) + 
  scale_shape_manual(values = c(21, 22, 21, 22, 24), 
                     name = "Site Type:",
                     labels = c("Alluvial Fan", "Colluvial Hollow",
                                "Hillslope Mineral Soil", "Ridgetop Mineral Soil",
                                "Valley Mineral Soil")) + 
  scale_color_manual(values = c("#D55E00", "#009E73","#202020", "#E69F00", "#56B4E9"),
                     name = "Site Type:",
                     labels = c("Alluvial Fan", "Colluvial Hollow",
                                "Hillslope Mineral Soil", "Ridgetop Mineral Soil",
                                "Valley Mineral Soil")) + 
  scale_fill_manual(values = c("#D55E0080","#009E7380","#20202080", "#E69F0080",
                               "#56B4E980"),
                    labels = c("Alluvial Fan", "Colluvial Hollow",
                                "Hillslope Mineral Soil", "Ridgetop Mineral Soil",
                                "Valley Mineral Soil"),
                    name = "Site Type:") + 
  background_grid() + panel_border() + scale_y_reverse() + labs(x = "Age (cal year BP)", y = "Depth (cm)", title = "Radiocarbon dated Charcoal")+ 
  theme(legend.position = "right") 
age_date_plot
save_plot("age_depth.png", age_date_plot, base_aspect_ratio = 1.5)
```


Saving just a copy of the two points
```{r}
date_point_ggplot <- plot_grid(ms_ggplot, ns_ggplot, labels = c("A.", "B."), nrow = 1, align = "h", rel_heights = c(1,1), rel_widths = c(12,12))
date_point_ggplot
save_plot("date_point.png", date_point_ggplot,ncol = 1, nrow = 2, base_aspect_ratio = 1.8, base_height = 4)
```

```{r}
dates_ggplot <- plot_grid(ms_ggplot, ns_ggplot, density_facet_plot, NULL, labels = c("A.", "B.", "C."))
dates_ggplot

# trying to fix alignment issue
# aligning top row (density plot) with bottom left (mineral soil)
plots <- align_plots(ms_ggplot, density_facet_plot, align = "v", axis = "l")
plots
# build bottom row
bottom_row <- plot_grid(plots[[1]], ns_ggplot, labels = c("B.", "C."), align = "h",axis = "b", rel_widths = c(1.5, 1.5))
plot_grid(plots[[2]], bottom_row, labels = c("A.", ""), ncol = 1, rel_heights = c(1.2, 1.2), rel_widths = c(0.5,1))
# nah I don't like how that looks

bottom_row <- plot_grid(ms_ggplot, ns_ggplot, labels = c("B.", "C."), align = "h", rel_widths = c(1.5,1.5))
date_facet_plot <- plot_grid(density_facet_plot, bottom_row, labels = c("A.", ""), ncol = 1, rel_heights = c(1.25, 1.5), rel_widths = c(0.95, 1))
date_facet_plot
# that's what I want but the y axis for the top and bottom aren't aligned...
# OH WELL FOR NOW
save_plot("date_facet_plot.png", date_facet_plot, nrow = 2, ncol = 2, base_aspect_ratio = 1.1)
```

Mmmm I think I'll try it with densities on the bottom, maybe the axes will align better in that position:
```{r}
# Switching order / fixing alignment issue
# aligning top row (soils) with bottom left (densities)
plots <- align_plots(ms_ggplot, density_facet_plot, align = "v", axis = "l")
# build bottom row
bottom_row <- plot_grid(plots[[1]], ns_ggplot, labels = c("B.", "C."), align = "h", axis = "b", rel_widths = c(1.5, 1.5))
plot_grid(plots[[2]], bottom_row, labels = c("A.", ""), ncol = 1, rel_heights = c(1.2, 1.2))
# nah I don't like how that looks
```


# Age Reversal
I also need to re-create the age-reversal graph I drew up in excel. I'm guessing the geom_point function should help with the plotting issues I've had, but I need to figure out how to cross-index the data appropriately.
I think if I start with each site_type category file, I can do them relatively quickly by hand and then add back together with an age_reversal / depth_reversal column for each. 
#### 5/6/18 - 
Just kidding, did it all in excel
```{r}
age_reverse <- read.csv('age_reverse.csv')

site_labels <- c("Alluvial Fan", "Colluvial Hollow","Hillslope Mineral Soil", "Ridgetop Mineral Soil", "Valley Mineral Soil")
site_colors <- c("#D55E00", "#009E73", "#202020", "#E69F00", "#56B4E9")
site_fill <- c("#D55E0080", "#009E7380", "#20202080", "#E69F0080", "#56B4E980")
  
age_reverse_ggplot <- ggplot(age_reverse, aes(x = age_reverse$depth_reversal, y = age_reverse$age_reversal, color = age_reverse$site_type, shape = age_reverse$site_type, fill = age_reverse$site_type)) + 
  geom_point(size = 4) + 
  scale_shape_manual(values = c(21,22,21, 22, 24),
                     name = "Site Type:",
                     labels = site_labels) +
  scale_color_manual(values = site_colors,
                     name = "Site Type:",
                     labels = site_labels) +
  scale_fill_manual(values = site_fill, name = "Site Type:",
                    labels = site_labels) +
  geom_hline(yintercept = 0) +
  background_grid( major = "xy") + panel_border() +
  xlab("Difference in depth in soil (cm)") + ylab("Difference in age (cal yr BP)") + 
  labs(title = "Age Reversal of Radiocarbon Dates") +
  annotate("text", x = c(88,85), y = c(-3800,4225), label = c("Age Reversal", "No Age Reversal")) 
age_reverse_ggplot
save_plot("age_reverse.png", age_reverse_ggplot, base_aspect_ratio = 1.8, base_height = 4)
```


```{r}
length(which(age_reverse$age_reversal < 0)) # age reversal
length(which(age_reverse$age_reversal > 0)) # no age reversal

```


# Linear Regression
Running a linear regression on age vs depth. 

```{r}
# running linear regression on MS dates
linearmod <- lm(formula = ms_dates$depth ~ ms_dates$median_cal_age, data = ms_dates)
print(linearmod)
summary(linearmod)
```

```{r}
plot(ms_dates$median_cal_age ~ ms_dates$depth, data = ms_dates)
abline(linearmod)
```

# comparing against norman
```{r}
dates
moderndates <- matrix(data = NA, ncol = 2, nrow = 6+length(norman))
moderndates[1:6,1] <- rep("soil")
moderndates[1:6,2] <- dates$median_cal_age[which(dates$median_cal_age <= 300)]
norman <- c(1764, 1766, 1774, 1808, 1834, 1850, 1873, 1883, 1917, 1936)
normanBP <- 2018 - norman
moderndates[7:16,1] <- rep("norman")
moderndates[7:16,2] <- normanBP

moderndates <- as.data.frame(moderndates)
colnames(moderndates) <- c("type", "date")
moderndates$date <- as.numeric(as.character(moderndates$date))
class(moderndates$date)

# scale_y_continuous(NULL, breaks = NULL)

overlap_ggplot <- ggplot(moderndates, aes(x = moderndates$date, color = moderndates$type, fill = moderndates$type)) +
  geom_dotplot( dotsize = 1.5, stackdir = "center") +
  scale_color_manual(values = c("#e34a33","#fdd0a2"),
                    name = "Data Type", 
                    labels = c("Tree-ring", "Soil") ) + 
  scale_fill_manual(values = c("#e34a3380","#fdd0a280"),
                    name = "Data Type", 
                    labels = c("Tree-ring", "Soil") ) +  
  scale_y_continuous(NULL, breaks = NULL) + 
  background_grid() +  panel_border() + labs(x = "Years BP", title = "Overlap between Tree-ring records and Soil Charcoal")
save_plot("overlap.png",overlap_ggplot, base_aspect_ratio = 1.8 )
```






